$\K$ of Kac---Moody groups, revisited

### 2017-06-17 paper.tex
Some fixes

\title{$\K$ of Kac---Moody groups, revisited}
\author {Sergey Sinchuk}
\address{Chebyshev laboratory, St. Petersburg State University, St. Petersburg, Russia}
\email {sinchukss at gmail.com}
\date {\today}

@@

In this note we describe a solution of the general problem, namely we describe an algorithm for computing $\K(A, F)$ for arbitrary $A$.
We also reprove some of the results of~\cite{MW} in a simplified form.
Our algorithm is, essentially, a variant of the well-known algorithm for computing Smith normal form of a matrix.
We also show~\cite[Conjecture 9.1]{MW} to be false as stated, and propose a way to corrected its statement.

\section{Proof of the main result}
Throughout this note $F$ denotes an arbitrary field and $A=(a_{ij})$ a generalized Cartan matrix of size $n$.
Denote by $M_A$ the subgroup of the free abelian group $\ZZ^n = \ZZ e_1 \oplus \ldots \oplus \ZZ e_n $ generated by all 
 linear combinations of the form $a_{ji} e_i - a_{ij} e_j$, $1\leq i<j\leq n$.
Set $G_A = \ZZ^n/M_A$.

@@

Now let $(c, p_c)$ be an $m\times n$ matrix with parity.
Define abelian group $L(p_c)$ as the direct sum $L_1 \oplus \ldots \oplus L_n$, where 
\[L_i=\left\{\def\arraystretch{1.2}%
  \begin{array}{@{}c@{\quad}l@{}}
     \K(F) & \text{if $p_c(i) = odd$,} \\    
     \Kmw(F) & \text{if $p_c(i) = even$}.
  \end{array}\right.\]
As before, we use notation $\{u, v\}_i$ for the Steinberg symbol (or cocycle) of $i$-th summand $L_i$.
We associate to $i$-th row $c_{i*}$ of $c$ the following element of $L(p_c)$:
\[r_{c,i}(u, v) = \{ u^{c_{i1}}, v\}_1 + \ldots + \{u^{c_{in}}, v\}_{n},\ \ u, v\in F^\times.\]
Define $J(c)$ to be the subgroup of $L(p_c)$ generated by $r_{c, i}(u, v)$ for $1\leq i\leq m$, $u, v\in F^\times.$
It is clear that $L(p_c)$ and $J({\widetilde{A}})$ coincide with $L$ and $J$ from the statement of~\cref{thm:mr}.

\begin{lemma} \label{lm:inv} The isomorphism class of the quotient $L(p_c)/J(c)$ is not changed
 by any of the above five elementary transformations of $(c, p_c)$. \end{lemma}
\begin{proof}
 %Assertion of the lemma is not immediately obvious only for operations of type \eqref{item:type4} and \eqref{item:type5}. 
 Let us verify the assertion of the lemma, for example, for the operation of type~\eqref{item:type5}
  (for other operations it is similar but easier).
 Assume that $c'$ is the matrix obtained from $c$ by applying operation of type~\eqref{item:type5}.

 To verify the assertion of the lemma it suffices to show that $J(c')$ is the image of $J(c)$ under an automorphism of $L(p_c)$.
 Speaking in terms of bases, it suffices to describe how the ``new'' symbols $\{u, v\}_i'$ corresponding to $c'$ 
  are expressed in terms of ``old'' symbols $\{u, v\}_i$ corresponding to $c$.

 By assumption, $c' = c \cdot t_{ij}(2k)$ for some $i$, $j$ satisfying $p_c(i)=odd$, $p_c(j)=even$.
 This operations corresponds to the following change of basis of $L(p_c)$:
 \[ \{u, v\}_{i}^{'} = \{u, v\}_i - 2k\{u, v\}_j;\ \ \{u, v\}_{j}^{'} = \{u, v\}_j.\]
 It remains to see that $\{u, v\}_{i}^{'}$ is a Steinberg symbol.
 Although $\{u, v\}_j$ is merely a cocycle, $2k\{u, v\}_j = \{u^{2k}, v\}_j$ is a symbol (cf.~\cite[Lemma~3.2(vii)]{MW}).
 Since $\{u, v\}_{i}^{'}$ is a linear combination of symbols, it is a symbol as well.
\end{proof}

\begin{lemma} \label{lm:normal-form} Any $m \times n$ integer matrix with parity $(c, p_c)$ can be transformed into a diagonal matrix $(d, p_d)$
  ($d = \mathrm{diag}(r_1, \ldots r_{\mathrm{min}(m, n)})$, $r_i | r_{i+1}$)
   by means of elementary operations of types~\eqref{item:type1}--\eqref{item:type5}.
\end{lemma}
\begin{proof} 
Recall that one of the working steps of the classical algorithm for finding Smith normal form involves
 addition of one of the two fixed columns (say, $i$-th and $j$-th ones) to another
 so that, unless it was zero from the start, the minimum of $|c_{ki}|$ and $|c_{kj}|$ decreases strictly 
 (here $k$ is some row number and $i$, $j$ are some column numbers selected by the algorithm on one if its steps).

To adapt the classical algorithm to the situation of matrices with parities it suffices to describe
 how the modified algorithm performs the same operation in the case when the selected columns have different parities.
Indeed, the our elementary operations differ from the usual ones only in that they do not allow adding odd multiples of odd columns to even columns.

@@

In the case $|c_{ki}| \geq |c_{kj}|$, write $c_{ki} + q c_{kj} = r$ for some $q$ and $r$, satisfying $|r| < |c_{kj}|$.
Using an operation of type~\eqref{item:type4} we add $q \cdot c_{*j}$ to $c_{*i}$ so that $\min(|c_{ki}|, |c_{kj}|)$ indeed decreases strictly.

In the case $|c_{ki}| < |c_{kj}|$, write $c_{kj} + q c_{ki} = r$ for some $q$ and $r$, satisfying $|r| < |c_{ki}|$.
If $q$ is even, we use an operation of type~\eqref{item:type5} to add $q \cdot c_{*i}$ to $c_{*j}$ and we are done again.
Otherwise, we can choose even $q'$ equal to either $q-1$ or $q+1$ so that one of the following is true:
\begin{itemize}
 \item $r\neq 0$ and $|c_{kj} + q' c_{ki}| < |c_{ki}|$;
 \item $r = 0$ and $c_{kj} + q' c_{ki} = c_{ki}$.
\end{itemize}
After adding $q' \cdot c_{*i}$ to $c_{*j}$ we are either done again (first case), or 
 need to further subtract $c_{*j}$ from $c_{*i}$ to get $c_{ki} = 0$ (second case).
\end{proof}

Once $(\widetilde{A}, p_{\widetilde{A}})$ is reduced to the diagonal form, the group $\K(A, F)$ can be calculated as follows.
Each zero diagonal coefficient $r_i$ produces one summand in $\K(A, F)$ which, depending on the parity of $i$-th column, is isomorphic to either $\K(F)$ or $\Kmw(F)$, whereas
 each nonzero coefficient $r_i \neq 0$ gives one summand $\K(F)/(r_i \K(F))$ or $\Kmw(F) / ((r_i/2) I)$.

Now suppose that columns of $A$ (and hence of $\widetilde{A}$) are ordered in such a way that
 each of the first $k$ columns contains and odd entry, and the remaining $n-k$ columns consist of even entries.
The algorithm described in~\cref{lm:normal-form} allows us to present
 $\widetilde{A}$ as $p \cdot d \cdot q $ for some $q \in \E(n, \ZZ)$, $p \in \E(n(n-1)/2, \ZZ)$, $d = \mathrm{diag}(r_1, \ldots r_n)$.

It is clear that $i$-th column of $d$ is odd if and only if coefficients $q_{ij}$ are even for $k+1\leq j \leq n$
 (to see this look at everything modulo $2$).
Alternatively, $i$-th column of $d$ is odd if and only if one of the coefficients $q'_{ji}$ of the inverse matrix $q^{-1}$ is odd for some $1\leq j\leq k$.

This means, that \cite[Conjecture~9.1]{MW} is true in the special case when Westaway's matrices $\nu$, $\mu$ (which equal our $q^t$, ${q^{-1}}^t$) 
 are obtained via the algorithm described above.
Notice, however, that this criterion is not applicable for $q$ obtained from {\it arbitrary} SNF decomposition $\widetilde{A} = p d q$.
Thus, as stated in~\cite{MW}, Conjecture~9.1 is false.

%A = 
%2, -2, -3, 0
%-2, 2, -2, -4
%-1, -2, 2, 0
%0, -4, 0, 2

\printbibliography
%\begin{thebibliography}{MMMM}
%\bibitem{Li}  {\sc B.~Liehl}, {On the group $SL_2$ over orders of arithmetic type}, J. Reine Angew. Math., 323, 153--171, (1981).
%\end{thebibliography}

>>>>

\title{$\K$ of Kac---Moody groups, revisited}
\author {Sergey Sinchuk}
%\address{Chebyshev laboratory, St. Petersburg State University, St. Petersburg, Russia}
%\email {sinchukss at gmail.com}
\date {\today}

@@

In this note we describe a solution of the general problem, namely we describe an algorithm which computues $\K(A, F)$ for arbitrary $A$.
We also reprove some of the results of~\cite{MW} in a simplified form.
Our algorithm is, essentially, a variant of the well-known algorithm for computing Smith normal form of a matrix.
%We also show~\cite[Conjecture 9.1]{MW} to be false as stated, and propose a way to corrected its statement.

\subsection{Proof of the main result}
Throughout this note $F$ denotes an arbitrary field and $A=(a_{ij})$ a generalized Cartan matrix of size $n$.
Denote by $M_A$ the subgroup of the free abelian group $\ZZ^n = \ZZ e_1 \oplus \ldots \oplus \ZZ e_n $ generated by all 
 linear combinations of the form $a_{ji} e_i - a_{ij} e_j$, $1\leq i<j\leq n$.
Set $G_A = \ZZ^n/M_A$.

@@

Now let $(c, p_c)$ be an $m\times n$ matrix with parity.
Define abelian group $L(p_c)$ as the direct sum $L_1 \oplus \ldots \oplus L_n$, where 
\[L_i=\left\{\def\arraystretch{1.2}%
  \begin{array}{@{}c@{\quad}l@{}}
     \K(F) & \text{if $p_c(i) = odd$,} \\    
     \Kmw(F) & \text{if $p_c(i) = even$}.
  \end{array}\right.\]
As before, we use notation $\{u, v\}_i$ for the Steinberg symbol (or cocycle) of $i$-th summand $L_i$.
Define $J(c)$ to be the subgroup of $L(p_c)$ generated by the following elements:
\[ \{ u^{c_{i1}}, v\}_1 + \ldots + \{u^{c_{in}}, v\}_{n},\ \ u, v\in F^\times, 1\leq i \leq m.\]
It is clear that $L(p_{\widetilde{A}})$ and $J({\widetilde{A}})$ coincide with $L$ and $J$ from the statement of~\cref{thm:mr}.

\begin{lemma} \label{lm:inv} The isomorphism class of the quotient $L(p_c)/J(c)$ is not changed
 by any of the above five elementary transformations of $(c, p_c)$. \end{lemma}
\begin{proof}
 %Assertion of the lemma is not immediately obvious only for operations of type \eqref{item:type4} and \eqref{item:type5}. 
 Let us verify the assertion of the lemma, for example, for the operation of type~\eqref{item:type5}
  (for other operations it is similar but easier).
 Assume that $c'$ is the matrix obtained from $c$ by applying some operation of type~\eqref{item:type5}.

 It suffices to show that $J(c')$ is the image of $J(c)$ under an automorphism of $L(p_c)$.
 Speaking in terms of bases, it suffices to describe how the ``new'' symbols $\{u, v\}_i'$ corresponding to $c'$ 
  are expressed in terms of ``old'' symbols $\{u, v\}_i$ corresponding to $c$.

 By assumption, $c' = c \cdot t_{ij}(2k)$ for some $i$, $j$ satisfying $p_c(i)=odd$, $p_c(j)=even$.
 This operations corresponds to the following change of basis of $L(p_c)$:
 \[ \{u, v\}_{i}^{'} = \{u, v\}_i - 2k\{u, v\}_j;\ \ \{u, v\}_{j}^{'} = \{u, v\}_j.\]
 It remains to see that $\{u, v\}_{i}^{'}$ is a Steinberg symbol.
 Indeed, although $\{u, v\}_j$ is merely a cocycle, $2k\{u, v\}_j = \{u^{2k}, v\}_j$ is a symbol (cf.~\cite[Lemma~3.2(vii)]{MW}).
 Since $\{u, v\}_{i}^{'}$ is a linear combination of symbols, it is a symbol as well.
\end{proof}

\begin{lemma} \label{lm:normal-form} Any $m \times n$ integer matrix with parity $(c, p_c)$ can be transformed into a diagonal matrix $(d, p_d)$
  ($d = \mathrm{diag}(r_1, \ldots r_{\mathrm{min}(m, n)})$, $r_i | r_{i+1}$)
   by means of elementary operations of types~\eqref{item:type1}--\eqref{item:type5}.
\end{lemma}
\begin{proof} 
Recall that the classical algorithm for finding Smith normal form of a matrix involves
 operation of addition of one of the two fixed columns (say, $i$-th and $j$-th ones) to another
 so that, unless it was zero from the start, the minimum of $|c_{ki}|$ and $|c_{kj}|$ decreases strictly 
 (here $k$ is some row number and $i$, $j$ are some column numbers selected by the algorithm on one if its steps).

To adapt the classical algorithm to our situation of matrices with parities it suffices to describe
 how the modified algorithm performs this operation in the case when the selected columns have different parities.
Indeed, our set of elementary operations differs from the usual one only in that
 we do not allow adding odd multiples of odd columns to even columns.

@@

%In the case $|c_{ki}| \geq |c_{kj}|$, write $c_{ki} + q c_{kj} = r$ for some $q$ and $r$, satisfying $|r| < |c_{kj}|$.
%Using an operation of type~\eqref{item:type4} we add $q \cdot c_{*j}$ to $c_{*i}$ so that $\min(|c_{ki}|, |c_{kj}|)$ decreases strictly.

We only need to consider the case $|c_{ki}| < |c_{kj}|$. Write $c_{kj} + q c_{ki} = r$ for some $q$ and $r$, satisfying $|r| < |c_{ki}|$.
If $q$ is even, add $q \cdot c_{*i}$ to $c_{*j}$ using an operation of type~\eqref{item:type5} and 
 notice that $\mathrm{min}(|c_{ki}|,|r|) < \mathrm{min}(|c_{ki}|,|c_{kj}|)$.
Otherwise, choose even $q'$ equal to either $q-1$ or $q+1$ so that one of the following is true:
\begin{itemize}
 \item $r\neq 0$ and $|c_{kj} + q' c_{ki}| < |c_{ki}|$;
 \item $r = 0$ and $c_{kj} + q' c_{ki} = c_{ki}$.
\end{itemize}
In the first case after adding $q' \cdot c_{*i}$ to $c_{*j}$ we obtain the required inequality.
In the second case we need to further subtract $c_{*j}$ from $c_{*i}$ to get $c_{ki} = 0$.
\end{proof}

After the canonical form $(d, p_d)$ of $(\widetilde{A}, p_{\widetilde{A}})$ is found, it becomes trivial to calculate the group $\K(A, F)$.
Each zero diagonal coefficient $r_i$ appearing in $d$ produces one summand in $\K(A, F)$ which, depending on $p_d(i)$, is isomorphic to $\K(F)$ or $\Kmw(F)$, whereas
 each nonzero coefficient $r_i \neq 0$ gives one summand $\K(F)/(r_i \K(F))$ or $\Kmw(F) / ((r_i/2) I)$.

Now suppose that columns of $A$ (and hence of $\widetilde{A}$) are ordered in such a way that
 each of the first $k$ columns contains and odd entry, and the remaining $n-k$ columns consist of even entries.
The algorithm given in~\cref{lm:normal-form} gives a decomposition
 $\widetilde{A} = p \cdot d \cdot q $ for some $q \in \E(n, \ZZ)$, $p \in \E(n(n-1)/2, \ZZ)$, $d = \mathrm{diag}(r_1, \ldots r_n)$.

Looking at the effect of operations \eqref{item:type2}--\eqref{item:type5} modulo $2$ one can note the following two properties of the matrix $q$:
\begin{itemize}
 \item $i$-th column of $d$ is odd if and only if coefficients $q_{ij}$ are even for $k+1\leq j \leq n$;
 \item $i$-th column of $d$ is odd if and only if one of the coefficients $q'_{ji}$ of the inverse matrix $q^{-1}$ is odd for some $1\leq j\leq k$.
\end{itemize}

In \cite{MW} M.~Westaway even conjectured that the above two properties could be used to read off from $q$ the parities of columns of $d$, and, thus, the type of summands appearing in $\K(A, F)$
 (Westaway's matrices $\nu$, $\mu$ appearing in the statement of \cite[Conjecture~9.1]{MW} correpond to our $q$ via $\nu = q^T$, $\mu = {q^{-1}}^T$).
While this characterization clearly works for the matrix $q$ obtained via the algorithm of~\cref{lm:normal-form},
 it may not work for $q$ obtained via the classical algorithm for finding SNF, which ignores columns' parities.
Thus, as stated, \cite[Conjecture~9.1]{MW} is false.
\begin{comment}
\begin{example}
To produce a counterexample, start from GCM 
 $A = \left(\begin{smallmatrix}
 2& -3& -2& 0\\
-1& 2& -2& 0\\
-2& -2& 2& -4\\
0& 0& -4& 2\end{smallmatrix}\right)$
  and write SNF decomposition of~\cref{lm:normal-form} for $\widetilde{A}$
\[\widetilde{A} = \left(\begin{smallmatrix}
-1& 3& 0& 0\\
-2& 0& 2& 0\\
0& 0& 0& 0\\
0& -2& 2& 0\\
0& 0& 0& 0\\
0& 0& -4& 4
\end{smallmatrix} \right) = \left(\begin{smallmatrix}
1& 0& 0& 0& 0& 0\\
2& 1& 0& 0& 0& 0\\
0& 0& 0& 0& 1& 0\\
0& 1& 0& 1& 0& 0\\
0& 0& 0& 0& 0& 1\\
0& -2& 1& 0& 0& 0\end{smallmatrix}\right) \cdot \mathrm{diag}(-1,2,4,4) \cdot \left(\begin{smallmatrix}
1& -3& 0& 0\\
0& -3& 1& 0\\
0& -3& 0& 1\\
0& 1& 0& 0\\ \end{smallmatrix}\right) = p \cdot d \cdot q.\]
Looking at the last 2 columns of $q$ we deduce that columns with $1$st and $4$th columns of $d$ are odd while $2$nd and $3$rd are even.
Now take $s = t_{43}(1) \in \E(4, \ZZ)$, $t = t_{43}(-1) \ in \E(6, \ZZ)$.
Notice that the central factor in the decomposition $\widetilde{A} = (p t) \cdot (t^{-1} d s^{-1}) \cdot (s q)$ is still 
 a still diagonal matrix but neither of the above two criterions is applicable to the matrix $sq$.
\end{example}
\end{comment}
\printbibliography
